---
title: "upa_analysis - draft"
author: "Peter Higgins"
date: "5/31/2022"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(here)
library(lubridate)
library(gt)
library(janitor)
library(ggdist)
library(ggthemes)
```

<!-- ## Upadacitinib Inpatients - Initial longitudinal plot -->

<!-- First pass at CRP values -->

<!-- ```{r data} -->
<!-- crp <- readxl::read_excel("upa_data2.xlsx") %>% -->
<!--   filter(test == "crp") %>%  -->
<!--   #filter(Dx == "uc") %>%  -->
<!--   mutate(patient = factor(patient)) %>%  -->
<!--   relocate(patient, Dx, upa_start_date, upa)  -->
<!-- ``` -->

<!-- ##  Plots -->

<!-- ```{r plot, echo=FALSE} -->
<!-- crp %>%  -->
<!--   group_by(patient) %>%  -->
<!--   ggplot() + -->
<!--   aes(x = date, y = 10*value, group = patient, color = patient) + -->
<!--   geom_point() + -->
<!--   geom_line() + -->
<!--   scale_color_manual(values = c("red", "blue", "green")) + -->
<!--   labs(y = "CRP in milligrams/Liter", -->
<!--        x = "Date", -->
<!--        title = "Inpatient Upadacitinib 30 mg po bid with IV steroids for ASUC") + -->
<!--   annotate("text", x = mdy_hms("May 27, 2022, 21:06:00"), y = 46, label= "Upa start 3", color = "blue"  ) + -->
<!--   annotate("text", x = mdy_hms("May 27, 2022, 14:04:00"), y = 238, label= "Upa start 1", color = "red") + -->
<!--   annotate("text", x = mdy_hms("May 29, 2022, 11:21:00"), y = 138, label= "Upa start 2", color = "blue") + -->
<!--   scale_y_continuous(breaks = seq(0, 250, 10), -->
<!--           expand = expansion(add = c(3,5))) + -->
<!--   geom_hline(yintercept =8, color = "black") + -->
<!--   annotate("text", x = mdy_hms("May 27, 2022, 15:00:00"), y = 10, label= "Upper Limit of Normal for CRP", color = "black") + -->
<!--   theme(legend.position = c(0.72, 0.796)) -->

<!-- ``` -->

## Data in 2 files
- all_hospitalized_uc.csv plus colectomy data - to be merged
- upa_ifx_tofa_hospitalized_uc.csv
Let's explore

all_hospitalized
- includes 10481 CRPs, 405 FCPs
- all are controls - no upa, tofa, ifx
- from 2012-02-11 to 2021-09-23
- 679 unique individuals, 198 went to colectomy

Now show average FCP by Day in the folks up to day 13 - most not in hospital very long. Many of these on Upa, as we started doing routine FCP on ASUC inpatients in the past year.

```{r}
all_hosp <- read_csv('all_hospitalized_uc.csv')
all_hosp |> distinct(cpi_num, .keep_all = TRUE) |> count()
# total 679 on CS, possibly some got IFX, some Cyclo, possibly a few tofa. Ended before Upa

all_hosp_colect <- readxl::read_excel('all_hospitalized_colectomy.xlsx') |> distinct(PatientID, .keep_all = TRUE)

all_hosp <- left_join(all_hosp, all_hosp_colect)

# how many colect

all_hosp_colect |> count()
# 198 colect/ total 679 on CS+
all_hosp |> 
  filter(test == "FCP") |> 
  select(cpi_num ,admit_date, 
         test_date, value) |> 
  mutate(day = as.numeric(test_date - admit_date + 1)) |> 
  group_by(day) |> 
  summarize(mean_fcp = mean(value),
            count = n()) |> 
  gt()

all_hosp |> 
  filter(test == "CRP") |> 
  select(cpi_num ,admit_date, 
         test_date, value) |> 
  mutate(day = as.numeric(test_date - admit_date + 1)) |> 
  group_by(day) |> 
  summarize(mean_crp = mean(value*10),
            count = n()) |> 
  print(n = 86) |> 
  gt()

```

Now let's look at advanced therapies
- 1890 lab values (Crp and FCP) for ifx, 368 for tofa, 243 for upa
- from 2014-01-06 to 2023-01-06
- 2455 CRPs, 46 FCPs
(note only recently routinely getting FCP, mostly in Upa)
- in 207 unique individuals
- among unique individuals
IFX - 61 colect, 95 not
tofa - 14 colect, 17 not
upa - 5 colect, 15 not

among previous group, treated with corticosteroids
CS - 198 colect/ 477 not, total 679 on CS

Mean CRP by Rx and colectomy at d3, d7,d14 with advanced Rx
(note rapid selection - people doing well go home, people doing badly go to colectomy, so N count goes down quickly, and FCP and CRP go up, as only the sick remain)

IFX no colect - d3 74.7, d7 33.5, d14 27.3
IFX eventual colect - d3 71, d7 59.7, d14 40.2
tofa no colect - d3 61.3, d7 32.7, d14 24.2
tofa eventual colect - d3 38.9, d7 20.5, d14 154
upa no colect - d3 41.2, d7 11.3, d14 79.2
upa eventual colect - d3 64.6, d7 51.4, d14 84

Full CRP table below

Then group by Rx and colectomy (0/1) and average daily value of CRP
(0-IFX means people who got IFX and no colectomy)

```{r}
adv_rx <- read_csv('upa_ifx_tofa_hospitalized_uc.csv')

adv_rx |> 
  filter(test == "FCP") |> 
  select(cpi_num,admit_date, cohort,
         test_date, value) |> 
  mutate(day = as.numeric(test_date - admit_date + 1)) |> 
  group_by(cohort, day) |> 
  summarize(mean_fcp = mean(value),
            count = n()) |> 
  gt()


adv_rx |> 
  filter(test == "CRP") |> 
  select(cpi_num,admit_date, cohort, colect,
         test_date, value) |> 
  mutate(day = as.numeric(test_date - admit_date + 1)) |> 
  group_by(colect, cohort, day) |> 
  summarize(mean_crp = mean(value*10),
            count = n()) |> 
  filter(day < 15) |> 
  gt()

adv_rx |> 
  distinct(cpi_num, ,.keep_all = TRUE) |> 
  tabyl(cohort, colect)
```

## When does colectomy occur?

Let's see when colectomies occur.
Are some treatments just delaying colectomy?

### Clean up and bind rows between controls and advRx first
```{r}
adv_rx2 <- adv_rx |> 
  select(cpi_num, cohort, admit_date, colect_date, colect) |> 
  rename(colectomy = colect) |> 
  replace_na(list(colectomy = 0)) |> 
  distinct(cpi_num, .keep_all = TRUE) |> 
  mutate(cpi_num = as.numeric(cpi_num)) 

all_hosp2 <- all_hosp |> 
  select(cpi_num, cohort, admit_date, Date, colectomy) |> 
  rename(colect_date = Date) |> 
  mutate(colect_date = as.Date( as.character(colect_date))) |> 
   distinct(cpi_num, .keep_all = TRUE) |> 
  replace_na(list(colectomy = 0))

all_colect <- bind_rows(adv_rx2, all_hosp2)
```

### Now set up plot

```{r}
all_colect |> 
  filter(colectomy == 1) |> 
  select(cohort, admit_date, colect_date) |> 
  mutate(day = as.numeric(colect_date - admit_date)) |>
  ggplot(aes(x = cohort, y = day, fill = fct_rev(cohort))) +
   stat_halfeye(adjust = 0.5, justification = -0.1, .width = 0) +
  geom_boxplot(
    width = 0.12,
    # removing outliers
    outlier.color = NA,
    alpha = 0.5
  ) +
  stat_dots(side = "left", 
            justification = 1.1, 
            binwidth = 5 ) +
  scale_y_continuous(limits = c(0,365)) +
  coord_flip() +
  theme(legend.position = c(0.8,0.85)) +
  labs(x = 'treatment', title = "When Does Colectomy Occur?", fill = "Treatment") +
  scale_x_discrete(labels = c("upa" = "Upa", "tofa" = "Tofa", "ifx" = "IFX", "ctls" = "CS")) +
  scale_fill_discrete(labels = c("Upa", "Tofa", "IFX", "Steroid"))
```

Looks like if we can catch people early enough with Upa, no colectomies after d39. Much bigger spread with other Advanced Rx

### Length of Followup for Upa?
Is it too short to be comparable?
Looks OK
```{r}
adv_rx |> 
  filter(cohort == "upa") |> 
  distinct(cpi_num, .keep_all = TRUE) |> 
  select(cohort, admit_date, colect_date) |> 
  mutate(cday = as.numeric(colect_date - admit_date)) |>
  mutate(fu = as.numeric(today() - admit_date)) |> 
  arrange(-fu)
```
range 130-409 days
median 307 days
One on 45 mg qd
One on 30 upa + stelara
All of the rest on 30 mg Upa

## How many colectomy by day 7 or 14?

```{r}
adv_rx |> 
  filter(colect == 1) |> 
  distinct(cpi_num, .keep_all = TRUE) |> 
  select(cohort, admit_date, colect_date) |> 
  mutate(day = as.numeric(colect_date - admit_date + 1)) |> 
  arrange(cohort, day) |> 
  print(n = Inf)

all_hosp2|> 
  filter(colectomy == 1) |> 
  distinct(cpi_num, .keep_all = TRUE) |> 
  select(cohort, admit_date, colect_date) |> 
  mutate(day = as.numeric(colect_date - admit_date + 1)) |> 
  arrange(cohort, day) |> 
  print(n = Inf)

```

*Colectomies*
IFX = 8/156 by day 7
IFX = 30/156 by day 14
tofa = 0/31 by day 7
tofa = 3/31 by day 14
upa = 2/20 by day 7
upa = 4/20 by day 14
CS - 113/697 by day 7
CS - 128/697 by day 14

Will need to pull more recent data to get
- CRP on CS patients
- FCP on CS patients
- more complete and accurate overall #s (note weirdness with CS and colectomy dates [some negative #s of days])