---
title: "upa_analysis - draft"
author: "Peter Higgins"
date: "5/31/2022"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(here)
library(lubridate)
library(gt)
library(janitor)
library(ggdist)
library(ggthemes)
```

<!-- ## Upadacitinib Inpatients - Initial longitudinal plot -->

<!-- First pass at CRP values -->

<!-- ```{r data} -->
<!-- crp <- readxl::read_excel("upa_data2.xlsx") %>% -->
<!--   filter(test == "crp") %>%  -->
<!--   #filter(Dx == "uc") %>%  -->
<!--   mutate(patient = factor(patient)) %>%  -->
<!--   relocate(patient, Dx, upa_start_date, upa)  -->
<!-- ``` -->

<!-- ##  Plots -->

<!-- ```{r plot, echo=FALSE} -->
<!-- crp %>%  -->
<!--   group_by(patient) %>%  -->
<!--   ggplot() + -->
<!--   aes(x = date, y = 10*value, group = patient, color = patient) + -->
<!--   geom_point() + -->
<!--   geom_line() + -->
<!--   scale_color_manual(values = c("red", "blue", "green")) + -->
<!--   labs(y = "CRP in milligrams/Liter", -->
<!--        x = "Date", -->
<!--        title = "Inpatient Upadacitinib 30 mg po bid with IV steroids for ASUC") + -->
<!--   annotate("text", x = mdy_hms("May 27, 2022, 21:06:00"), y = 46, label= "Upa start 3", color = "blue"  ) + -->
<!--   annotate("text", x = mdy_hms("May 27, 2022, 14:04:00"), y = 238, label= "Upa start 1", color = "red") + -->
<!--   annotate("text", x = mdy_hms("May 29, 2022, 11:21:00"), y = 138, label= "Upa start 2", color = "blue") + -->
<!--   scale_y_continuous(breaks = seq(0, 250, 10), -->
<!--           expand = expansion(add = c(3,5))) + -->
<!--   geom_hline(yintercept =8, color = "black") + -->
<!--   annotate("text", x = mdy_hms("May 27, 2022, 15:00:00"), y = 10, label= "Upper Limit of Normal for CRP", color = "black") + -->
<!--   theme(legend.position = c(0.72, 0.796)) -->

<!-- ``` -->

## Let's explore asuc
1. How many admissions - 1360
2. How many to colectomy - 513
3. ifx 110 colect, 156 not
4. cyclo 4 colect, 12 not
5. tofa 37 colect, 52 not
6. upa 7 colect, 15 not
7. steroids only 360 colect, 617 not


```{r}
asuc <- read_csv("Berinstein_UC_2014-2023.csv") |> 
  mutate(colect = coalesce(colect, 0)) |>
  mutate(exists = 1) |> 
  pivot_wider(names_from = cohort, values_from = exists, values_fill = 0) |> 
  mutate(lab_value = as.numeric(lab_value))

# how many unique individual admissions
asuc |> distinct(cpi_num, admit_date) |>
  count()
# 1360
# All comers - 
# how many went to colectomy? #513 yes, 847 no
asuc |> distinct(cpi_num, admit_date, .keep_all = TRUE) |>
   count(colect)

# by group
asuc |> distinct(cpi_num, admit_date, .keep_all = TRUE) |>
  filter(ifx == 1) |> 
   count(colect)
# ifx 110 colect, 156 not
asuc |> distinct(cpi_num, admit_date, .keep_all = TRUE) |>
  filter(cyclo == 1) |> 
   count(colect)
# cyclo 4 colect, 12 not
asuc |> distinct(cpi_num, admit_date, .keep_all = TRUE) |>
  filter(tofa == 1) |> 
   count(colect)
# tofa 37 colect, 52 not
asuc |> distinct(cpi_num, admit_date, .keep_all = TRUE) |>
  filter(upa == 1) |> 
   count(colect)
# upa 7 colect, 15 not

asuc |> distinct(cpi_num, admit_date, .keep_all = TRUE) |>
  filter(steroid == 1 & ifx == 0 & cyclo == 0 & tofa == 0 & upa == 0) |> 
   count(colect)
# steroids only 360 colect, 617 not
```


Let's look at individual therapies for FCP and CRP

Steroids only
```{r}
# FCP values, steroid only
asuc |> 
  filter(test == "FCP") |> 
   distinct(cpi_num, admit_date, .keep_all = TRUE) |> 
    filter(steroid == 1 & ifx == 0 & cyclo == 0 & tofa == 0 & upa == 0) |> 
  select(cpi_num ,admit_date, 
         test_date, lab_value) |> 
  mutate(day = as.numeric(test_date - admit_date + 1)) |> 
  group_by(day) |> 
  filter(day <9) |> 
  summarize(mean_fcp = mean(lab_value, na.rm = TRUE),
            count = n()) |> 
  gt() |> 
  tab_header(title = "FCP with steroid monotherapy")

# CRP values, steroid only
asuc |> 
  filter(test == "CRP") |>    distinct(cpi_num, admit_date, .keep_all = TRUE) |> 
    filter(steroid == 1 & ifx == 0 & cyclo == 0 & tofa == 0 & upa == 0) |> 
  select(cpi_num ,admit_date, 
         test_date, lab_value) |> 
  mutate(day = as.numeric(test_date - admit_date + 1)) |> 
  group_by(day) |> 
  filter(day < 9) |> 
  summarize(mean_crp = mean(lab_value*10, na.rm = TRUE),
            count = n()) |> 
  print(n = 86) |> 
  gt() |> 
  tab_header(title = "CRP with steroid monotherapy")

```



IFX treatment
```{r}
# FCP values, ifx only
asuc |> 
  filter(test == "FCP") |> 
   distinct(cpi_num, admit_date, .keep_all = TRUE) |> 
    filter(ifx == 1 ) |> 
  select(cpi_num ,admit_date, 
         test_date, lab_value) |> 
  mutate(day = as.numeric(test_date - admit_date + 1)) |> 
  group_by(day) |> 
  filter(day <9) |> 
  summarize(mean_fcp = mean(lab_value, na.rm = TRUE),
            count = n()) |> 
  gt() |> 
  tab_header(title = "FCP with IFX therapy")

# CRP values, ifx only
asuc |> 
  filter(test == "CRP") |>    distinct(cpi_num, admit_date, .keep_all = TRUE) |> 
    filter(ifx == 1 ) |>  
  select(cpi_num ,admit_date, 
         test_date, lab_value) |> 
  mutate(day = as.numeric(test_date - admit_date + 1)) |> 
  group_by(day) |> 
  filter(day < 9) |> 
  summarize(mean_crp = mean(lab_value*10, na.rm = TRUE),
            count = n()) |> 
  print(n = 86) |> 
  gt() |> 
  tab_header(title = "CRP with IFX therapy")

```

Cyclo treatment
```{r}
# FCP values, cyclo only
asuc |> 
  filter(test == "FCP") |> 
   distinct(cpi_num, admit_date, .keep_all = TRUE) |> 
    filter(cyclo == 1 ) |> 
  select(cpi_num ,admit_date, 
         test_date, lab_value) |> 
  mutate(day = as.numeric(test_date - admit_date + 1)) |> 
  group_by(day) |> 
  filter(day <9) |> 
  summarize(mean_fcp = mean(lab_value, na.rm = TRUE),
            count = n()) |> 
  gt() |> 
  tab_header(title = "FCP with Cyclo therapy")

# CRP values, cyclo only
asuc |> 
  filter(test == "CRP") |>    distinct(cpi_num, admit_date, .keep_all = TRUE) |> 
    filter(cyclo == 1 ) |>  
  select(cpi_num ,admit_date, 
         test_date, lab_value) |> 
  mutate(day = as.numeric(test_date - admit_date + 1)) |> 
  group_by(day) |> 
  filter(day < 9) |> 
  summarize(mean_crp = mean(lab_value*10, na.rm = TRUE),
            count = n()) |> 
  print(n = 86) |> 
  gt() |> 
  tab_header(title = "CRP with Cyclo therapy")

```


Tofa treatment
```{r}
# FCP values, Tofa only
asuc |> 
  filter(test == "FCP") |> 
   distinct(cpi_num, admit_date, .keep_all = TRUE) |> 
    filter(tofa == 1 ) |> 
  select(cpi_num ,admit_date, 
         test_date, lab_value) |> 
  mutate(day = as.numeric(test_date - admit_date + 1)) |> 
  group_by(day) |> 
  filter(day <9) |> 
  summarize(mean_fcp = mean(lab_value, na.rm = TRUE),
            count = n()) |> 
  gt() |> 
  tab_header(title = "FCP with Tofa therapy")

# CRP values, tofa only
asuc |> 
  filter(test == "CRP") |>    distinct(cpi_num, admit_date, .keep_all = TRUE) |> 
    filter(tofa == 1 ) |>  
  select(cpi_num ,admit_date, 
         test_date, lab_value) |> 
  mutate(day = as.numeric(test_date - admit_date + 1)) |> 
  group_by(day) |> 
  filter(day < 9) |> 
  summarize(mean_crp = mean(lab_value*10, na.rm = TRUE),
            count = n()) |> 
  print(n = 86) |> 
  gt() |> 
  tab_header(title = "CRP with Tofa therapy")

```



Upa treatment
```{r}
# FCP values, Upa only
asuc |> 
  filter(test == "FCP") |> 
   distinct(cpi_num, admit_date, .keep_all = TRUE) |> 
    filter(upa == 1 ) |> 
  select(cpi_num ,admit_date, 
         test_date, lab_value) |> 
  mutate(day = as.numeric(test_date - admit_date + 1)) |> 
  group_by(day) |> 
  filter(day <9) |> 
  summarize(mean_fcp = mean(lab_value, na.rm = TRUE),
            count = n()) |> 
  gt() |> 
  tab_header(title = "FCP with Upa therapy")

# CRP values, Upa only
asuc |> 
  filter(test == "CRP") |>    distinct(cpi_num, admit_date, .keep_all = TRUE) |> 
    filter(upa == 1 ) |>  
  select(cpi_num ,admit_date, 
         test_date, lab_value) |> 
  mutate(day = as.numeric(test_date - admit_date + 1)) |> 
  group_by(day) |> 
  filter(day < 9) |> 
  summarize(mean_crp = mean(lab_value*10, na.rm = TRUE),
            count = n()) |> 
  print(n = 86) |> 
  gt() |> 
  tab_header(title = "CRP with Upa therapy")

```


## When does colectomy occur?

Let's see when colectomies occur.
Are some treatments just delaying colectomy?



## Retry with Beeswarm
```{r}
library(ggbeeswarm)
asuc |> 
  filter(colect == 1) |> 
  mutate(cohort = factor(case_when(
    ifx == 1 ~ "IFX",
    tofa == 1 ~ "Tofa",
    upa == 1 ~ "Upa",
    .default = "Steroids"
  ), levels = rev(c("Upa", "Tofa", "Cyclo", "Steroids", "IFX")))
  )  |> distinct(cpi_num, .keep_all = TRUE) |> 
  ggplot(aes(x = cohort, 
             y = time_to_colect,
             color = fct_rev(cohort))) +
  geom_quasirandom(method = "tukey") +
  scale_y_continuous(limits = c(0,730)) +
  coord_flip() +
  labs(x = 'treatment', title = "When Does Colectomy Occur?", y = "Days to Colectomy", color = "Treatment",
       subtitle = "Are Steroids and IFX Just Delaying Colectomy in Many?") +
      theme_linedraw()  + theme(legend.position = c(0.8,0.85))
```

Looks like if we can catch people early enough with Upa, no colectomies after d39. Much bigger spread with other Advanced Rx

### Length of Followup for Upa?
Is it too short to be comparable?
Looks OK

```{r}
asuc |> 
  mutate(cohort = factor(case_when(
    ifx == 1 ~ "IFX",
    tofa == 1 ~ "Tofa",
    upa == 1 ~ "Upa",
    .default = "Steroids"
  ), levels = rev(c("Upa", "Tofa", "Cyclo", "Steroids", "IFX")))) |> 
  filter(cohort == "Upa") |> 
  distinct(cpi_num, .keep_all = TRUE) |> 
  select(cohort, admit_date, colect_date) |> 
  mutate(cday = as.numeric(colect_date - admit_date)) |>
  mutate(fu = as.numeric(today() - admit_date)) |> 
  arrange(-fu)
```
range 29-367 days
median 251 days
One on 45 mg qd
One on 30 upa + stelara
All of the rest on 30 mg Upa

## How many colectomy by day 7 or 14?

```{r}
# count all treated by Rx
asuc |> 
   mutate(cohort = factor(case_when(
    ifx == 1 ~ "IFX",
    tofa == 1 ~ "Tofa",
    upa == 1 ~ "Upa",
    .default = "Steroids"
  ), levels = rev(c("Upa", "Tofa", "Cyclo", "Steroids", "IFX")))) |> 
  distinct(cpi_num, .keep_all = TRUE) |> 
  group_by(cohort) |> 
  count()

# count all colectomies by Rx
asuc |> 
  filter(colect == 1) |> 
   mutate(cohort = factor(case_when(
    ifx == 1 ~ "IFX",
    tofa == 1 ~ "Tofa",
    upa == 1 ~ "Upa",
    .default = "Steroids"
  ), levels = rev(c("Upa", "Tofa", "Cyclo", "Steroids", "IFX")))) |> 
  distinct(cpi_num, .keep_all = TRUE) |> 
  group_by(cohort) |> 
  count()

asuc |> 
  filter(colect == 1) |> 
   mutate(cohort = factor(case_when(
    ifx == 1 ~ "IFX",
    tofa == 1 ~ "Tofa",
    upa == 1 ~ "Upa",
    .default = "Steroids"
  ), levels = rev(c("Upa", "Tofa", "Cyclo", "Steroids", "IFX")))) |> 
  distinct(cpi_num, .keep_all = TRUE) |> 
  select(cohort, time_to_colect) |>
  arrange(cohort, time_to_colect) |> 
  print(n = Inf)

# Note a bunch of NEGATIVE time to colect

```

*Colectomies*
IFX = 29+3?/68 by day 7 - 186 treated <br>
IFX = 41+3?/68 by day 14 - 186 treated <br>
tofa = 6+1?/10 by day 7 - 41 treated <br>
tofa = 6+1?/10 by day 14 - 41 treated <br>
upa = 2/3 by day 7 - 11 treated <br>
upa = 3/3 by day 14 - 11 treated <br>
CS - 83+?/171 by day 7 - 607 treated <br>
CS - 93+?/171 by day 14 - 607 treated <br>

Will need to check the data to get

- more complete and accurate overall #s (note weirdness with colectomy dates [some negative #s of days])
- also upa total seems low?
- tofa total may be low